---
title: "Code assembly Cystic Fibrosis food allergens"
author: 
  - Marc Emmenegger^[Division of Medical Immunology, University Hospital Basel, marc.emmenegger@usb.ch], 
date: "2026-07-05"
output: 
  html_document:
    number_sections: TRUE
    keep_md: yes
    css: style.css
    code_folding: "hide"
documentclass: article
fontsize: 11pt
linkcolor: blue
mainfont: Arial
toc: true
toc_float: true
toc_depth: 6
mathjax: null
css: doc.css
editor_options: 
  chunk_output_type: console
---

# GENERAL COMMENT TO THIS RMARKDOWN FILE

This Rmarkdown file contains code written in R utilised for parts of the analyses contained in the manuscript.

# LIBRARIES

```{r LIBRARIES, message=FALSE}

# Core tidyverse & data manipulation
library(tidyverse)
library(dplyr)
library(tidyr)
library(purrr)
library(magrittr)
library(stringr)
library(stringi)
library(forcats)
library(janitor)
library(broom)
library(lubridate)

# Data import/export
library(openxlsx)

# Visualization
library(ggplot2)
library(ggpubr)
library(cowplot)
library(ggridges)
library(ggh4x)
library(gplots)
library(RColorBrewer)
library(corrplot)
library(viridis)
library(ggsci)
library(hrbrthemes)
library(heatmaply)
library(plotrix)
library(table1)
library(DT)
library(knitr)

# Clustering, dimension reduction & multivariate analysis
library(cluster)
library(factoextra)
library(FactoMineR)
library(umap)
library(Rtsne)
library(bigmds)
library(parallelDist)
library(vegan)
library(dendextend)
library(ggdendro)
library(Boruta)
library(circlize)

# Statistics & modeling
library(MASS)
library(mgcv)
library(lme4)
library(rstanarm)
library(drc)
library(psych)
library(Hmisc)
library(MESS)
library(splines)
library(PupillometryR)
library(fmsb)
library(gapminder)
library(Lahman)

# Base R
require(graphics)

# Global options
options(mc.cores = parallel::detectCores())
theme_set(theme_bw())

```

# ANALYSES PERTAINING TO FIGURE 1 AND S1

```{r ANALYSES PERTAINING TO FIGURE 1 AND S1, message=FALSE}

################################################################################
# Fig. 1C
################################################################################

ggplot(data=departments_data_count_df, aes(x=reorder(Var1, -Freq), y=Freq, fill=reorder(Var1, -Freq))) +
  geom_bar(stat="identity") +
  labs(y="Number of occurrences", x='Departments') +
  coord_flip()

################################################################################
# Fig. 1D
################################################################################

qplot(numbers_df$numbers, geom="histogram", ylab = 'Distinct ICD-10 codes', xlab = 'Number of samples per ICD-10 code', binwidth = 10)

################################################################################
# Fig. 1E
################################################################################

plot(Arah2_histo$pEC50)

################################################################################
# Fig. S1B
################################################################################

ggqqplot(data$posterior)

ggplot(Arah2_histo, aes(x = pEC50)) + 
  geom_histogram(aes(y =..density..),
                 breaks = seq(0, 4, by = 0.1), 
                 colour = "black", 
                 fill = "pink") +
                 stat_function(fun = dnorm, args = list(mean = mean(Arah2_histo$pEC50), sd = sd(Arah2_histo$pEC50)))

################################################################################
# Fig. S1C
################################################################################

#### Ara h 2 and PrP

ggplot(overlap_data, aes(x=Arah2, y=PrP)) +
  geom_point()

#### Ara h 2 and CGRP

ggplot(overlap_data, aes(x=Arah2, y=CGRP)) +
  geom_point()

#### Distributions

ggplot(overlap_data_joy_Arah2, aes(x = pEC50, y = Label, fill = Label)) +
  geom_density_ridges(
    aes(point_color = Label, point_fill = Label, point_shape = Class_hits, point_size = Size),
    alpha = .2, point_alpha = 1, jittered_points = TRUE
  ) +
  scale_point_color_hue(l = 40) +
  ggtitle('Ridge plot with shape and size different for hits for Ara h 2 - data from Ara h 2') +
  xlab('p(EC50)') +
  ylab('Antigen') +
  scale_x_continuous(limits = c(0, 4))
  
ggplot(overlap_data_joy_PrP, aes(x = pEC50, y = Label, fill = Label)) +
  geom_density_ridges(
    aes(point_color = Label, point_fill = Label, point_shape = Class_hits, point_size = Size),
    alpha = .2, point_alpha = 1, jittered_points = TRUE
  ) +
  scale_point_color_hue(l = 40) +
  ggtitle('Ridge plot with shape and size different for hits for Ara h 2 - data from PrP') +
  xlab('p(EC50)') +
  ylab('Antigen') +
  scale_x_continuous(limits = c(0, 4))

ggplot(overlap_data_joy_CGRP, aes(x = pEC50, y = Label, fill = Label)) +
  geom_density_ridges(
    aes(point_color = Label, point_fill = Label, point_shape = Class_hits, point_size = Size),
    alpha = .2, point_alpha = 1, jittered_points = TRUE
  ) +
  scale_point_color_hue(l = 40) +
  ggtitle('Ridge plot with shape and size different for hits for Ara h 2 - data from CGRP') +
  xlab('p(EC50)') +
  ylab('Antigen') +
  scale_x_continuous(limits = c(0, 4))

################################################################################
# Fig. S1D
################################################################################

rnames <- data[,1]
print(rnames)
rownames(data) <- rnames

scatter_data <- as.data.frame(data[,2:3])

scatter_data_lm1 <- scatter_data$First ~ scatter_data$Second

ggplot(scatter_data, aes(x=Second, y=First)) +
  geom_point(size=2, shape=1, color = 'black') +
  geom_smooth(method="lm", se=TRUE, fullrange=FALSE, level=0.95, color='darkred', fill='blue') +
  coord_cartesian(ylim = c(0, 4), xlim = c(0,4)) 

```

# ANALYSES PERTAINING TO FIGURE 2

```{r ANALYSES PERTAINING TO FIGURE 2, message=FALSE}

################################################################################
# Figs. 2A and 2B
################################################################################

ggplot(data_sex, aes(fill=Sex, y=Number_percent, x=Condition)) + 
  geom_bar(position="stack", stat="identity")

ggplot(data_age_points, aes(fill=Sex, y=Age, x=Condition)) + 
  geom_boxplot()+
  geom_dotplot(position="dodge", stat="identity", binaxis='y', stackdir = 'center', dotsize = 0.1, binwidth = 1)

################################################################################
# Fig. 2C
################################################################################

ggplot(data=df_all_vs_seropositives_dep_aligned, aes(x=reorder(Var1, -Freq_condition), y=Freq_percent, fill=Condition)) +
  geom_bar(stat="identity", position="identity", alpha = .2) +
  labs(y="Percent of condition", x='Departments') +
  scale_fill_manual(values=c("#34baeb", "#9e34eb")) +
  coord_flip()

################################################################################
# Figs. 2D and 2E
################################################################################

#### UMAP for dataset with INCLUSION of male/female sex

groups <- as.factor(data$Class[1:25597])
label <- unique(groups)
print(groups)
print(label)

custom.config = umap.defaults
custom.config$metric = 'cosine'
custom.config

set.seed(42)
umap_04 <- umap::umap(data.matrix(data[c(3:1534, 1536)]), metric = 'cosine')

df_umap_04 <- data.frame(x = umap_04$layout[,1],
                 y = umap_04$layout[,2],
                 color = groups)

ggplot(df_umap_04, aes(x, y, color = groups)) +
  geom_point(size = 2, stroke = 0) +
  theme_bw() + 
  ylab("UMAP Dimension 2") + 
  xlab("UMAP Dimension 1") + 
  scale_color_manual(values=c("#6A8ED2", "red")) +
  labs(colour="Hits") +
  ggtitle('Cosine UMAP of all patients screened: ICD-10 and sex')

#### PCA for dataset with INCLUSION of male/female sex

mat_data <- data.matrix(data[c(3:1534, 1536)])
print(mat_data)
rownames(mat_data) <- rnames          
print(rownames(mat_data))

groups <- as.factor(data$Class[1:25597])
print(groups)

set.seed(42)
mat_data.pca <- prcomp(mat_data, center = TRUE)
print(mat_data.pca)

fviz_eig(mat_data.pca)

summary(mat_data.pca)

fviz_pca_ind(mat_data.pca,
             col.ind = groups,
             palette = c("#6A8ED2",  "red"), 
             addEllipses = FALSE,
             legend.title = "Hits", 
             repel = TRUE, 
             geom.ind = c("point"),
             title = 'PCA of all patients screened: ICD-10 and sex'
)

################################################################################
# Fig. 2G
################################################################################

logit = function(x) log(x/(1-x))
inv.logit = function(x) exp(x)/(1+exp(x))
qsum = function(x) c(quantile(x,probs=c(0.5,0.025,0.975)))

# The approach is simple: 
# (1) We use logit transformed EC50 values. 
# (2) We conduct a gaussian regression. 
# (3) We exponentiate the resulting to obtain the OR.
# Compared to standard logistic regression, we carry out the logit-transform beforehand and do not
# run a binomial regression model with logit link.
# Instead, we run a Gaussian regression on logit-transformed data.

#### 1.

rnames <- data[,1]
print(rnames)

#### 2.

# We (1) exponentiate the pEC50 to get the EC50 and then (2) negatively logit transform the EC50 values.
# The highest p(EC50) values therefore correspond to the highest -logit(EC50) values.

icd10_logit =  data %>%
  dplyr::mutate(age=(age-mean(age))/sd(age)) %>%
  dplyr::mutate(posterior=10^(-posterior)) %>%
  dplyr::mutate(posterior=ifelse(posterior==0,0+0.0260638,posterior)) %>%
  dplyr::mutate(posterior=ifelse(posterior==1,1-0.0260638,posterior)) %>% #We scale between 0-1, so that we reach the value of the highest sample that is not 1
  dplyr::mutate(posterior=-logit(posterior)) %>%
  dplyr::select(-patient_id) 

#### 3.

# normal prior
m1_lin = stan_glm(posterior~.,
                  family=gaussian(),
                  prior=normal(0,10),
                  prior_intercept = normal(0,10),
                  data=icd10_logit)
summary(m1_lin)
s1_lin = summary(m1_lin,probs = c(0.025,.5,0.975)) %>%
  as.data.frame() %>%
  rownames_to_column() %>% 
  as_tibble() %>%
  filter(!(rowname %in% c("log-posterior","mean_PPD","(Intercept)","sigma"))) %>%
  mutate(rowname=gsub("TRUE","",rowname),
         OR=exp(`50%`),
         OR_low=exp(`2.5%`),
         OR_high=exp(`97.5%`),
         type="Uninformative prior") %>%
  arrange(-abs(`50%`))


# lasso prior with 1 df
m1_lasso = stan_glm(posterior~.,
                    family=gaussian(),
                    prior=lasso(df=1),
                    data=icd10_logit)
summary(m1_lasso)
s1_lasso = summary(m1_lasso,probs = c(0.025,.5,0.975)) %>%
  as.data.frame() %>%
  rownames_to_column() %>% 
  as_tibble() %>%
  filter(!(rowname %in% c("log-posterior","mean_PPD","(Intercept)","sigma"))) %>%
  mutate(rowname=gsub("TRUE","",rowname),
         OR=exp(`50%`),
         OR_low=exp(`2.5%`),
         OR_high=exp(`97.5%`),
         type="LASSO") %>%
  arrange(-abs(`50%`))

# horseshoe prior (Piironen and Vehtari (2017))  
globalscale = (19/570)/sqrt(589)

m1_hs = stan_glm(posterior~.,
                 family=gaussian(),
                 prior=hs(global_scale=globalscale),
                 data=icd10_logit)
summary(m1_hs)
s1_hs = summary(m1_hs,probs = c(0.025,.5,0.975)) %>%
  as.data.frame() %>%
  rownames_to_column() %>% 
  as_tibble() %>%
  filter(!(rowname %in% c("log-posterior","mean_PPD","(Intercept)","sigma"))) %>%
  mutate(rowname=gsub("TRUE","",rowname),
         OR=exp(`50%`),
         OR_low=exp(`2.5%`),
         OR_high=exp(`97.5%`),
         type="Regularized horseshoe") %>%
  arrange(-abs(`50%`))

post_icd10 = bind_rows(s1_lasso,s1_hs,s1_lin) %>%
  mutate(rowname=factor(rowname,levels=s1_hs$rowname[order(s1_hs$`50%`)]),
         type=factor(type,levels=c("Uninformative prior","LASSO","Regularized horseshoe")))
saveRDS(post_icd10,file="post_icd10.rds")

#### 4.

post_icd10 = readRDS("post_icd10.rds")
ranks = filter(post_icd10,type=="Regularized horseshoe") %>%
  arrange(-`50%`) %>%
  mutate(rank=row_number()) %>%
  mutate(ICD10=rowname) %>%
  dplyr::select(ICD10,rank)

tibble_post_icd10 <- post_icd10 %>%
  mutate(OR=paste0(sprintf("%.2f",OR)," (",sprintf("%.2f",OR_low),"-",sprintf("%.2f",OR_high),")")) %>%
  dplyr::select(ICD10=rowname,OR,type) %>%
  pivot_wider(names_from=type,values_from=OR) %>%
  left_join(ranks) %>%
  arrange(rank) %>% 
  filter(rank %in% c(1:10,(max(rank)-9):max(rank))) %>%
  dplyr::select(rank,ICD10,`Uninformative prior`,LASSO,`Regularized horseshoe`)
write.csv(tibble_post_icd10,"tibble_post_icd10_ranktable.csv")

post_icd10 %>%
  mutate(type=factor(type,labels=c("Unregularized","LASSO","Regularized Horseshoe"))) %>%
  ggplot() +
  geom_pointrange(aes(x=rowname,y=OR,ymin=OR_low,ymax=OR_high,colour=type)) +
  geom_hline(yintercept=1,linetype=2) +
  scale_colour_discrete(guide='none') +
  scale_y_log10() +
  coord_flip() +
  facet_wrap(~type,ncol=3) +
  theme(axis.text.y = element_text(size = 1)) +
  labs(y="Odds ratio (95% credible interval)",x="ICD 10 code, sex, or age",title="Association between ICD-10 codes and pEC50 values against nAra h 2.")

write.csv(post_icd10,"post_icd10_ranktable.csv")

post_icd10_top10 %>%
  mutate(type) %>%
  ggplot() +
  geom_pointrange(aes(reorder(x=rowname, -rank),y=OR,ymin=OR_low,ymax=OR_high,colour=type)) +
  geom_hline(yintercept=1,linetype=2) +
  scale_colour_discrete(guide='none') +
  scale_y_log10() +
  coord_flip() +
  facet_wrap(~type,ncol=3) +
  theme(axis.text.y = element_text(size = 10)) +
  labs(y="Odds ratio (95% credible interval)",x="ICD-10 code, sex, or age",title="Association between ICD-10 codes and -logit(EC50) values against nAra h 2.")

################################################################################
# Fig. 2I
################################################################################

logit = function(x) log(x/(1-x))
inv.logit = function(x) exp(x)/(1+exp(x))
qsum = function(x) c(quantile(x,probs=c(0.5,0.025,0.975)))

medication_logit =  data %>%
  dplyr::mutate(age=(age-mean(age))/sd(age)) %>%
  dplyr::mutate(posterior=10^(-posterior)) %>%
  dplyr::mutate(posterior=ifelse(posterior==0,0+0.0260638,posterior)) %>%
  dplyr::mutate(posterior=ifelse(posterior==1,1-0.0260638,posterior)) %>% #We scale between 0-1, so that we reach the value of the highest sample that is not 1
  dplyr::mutate(posterior=-logit(posterior)) %>%
  dplyr::select(-patient_id) 

qplot(medication_logit$posterior, geom="histogram", ylab = 'Number of occurrences', xlab = '-logit(EC50)', binwidth = 0.05)
ggqqplot(medication_logit$posterior) #We observe some very negative values that result from the -logit transform... 

# normal prior
m1_lin = stan_glm(posterior~.,
                  family=gaussian(),
                  prior=normal(0,10),
                  prior_intercept = normal(0,10),
                  data=medication_logit)
summary(m1_lin)
s1_lin = summary(m1_lin,probs = c(0.025,.5,0.975)) %>%
  as.data.frame() %>%
  rownames_to_column() %>% 
  as_tibble() %>%
  filter(!(rowname %in% c("log-posterior","mean_PPD","(Intercept)","sigma"))) %>%
  mutate(rowname=gsub("TRUE","",rowname),
         OR=exp(`50%`),
         OR_low=exp(`2.5%`),
         OR_high=exp(`97.5%`),
         type="Uninformative prior") %>%
  arrange(-abs(`50%`))

# lasso prior with 1 df
m1_lasso = stan_glm(posterior~.,
                    family=gaussian(),
                    prior=lasso(df=1),
                    data=medication_logit)
summary(m1_lasso)
s1_lasso = summary(m1_lasso,probs = c(0.025,.5,0.975)) %>%
  as.data.frame() %>%
  rownames_to_column() %>% 
  as_tibble() %>%
  filter(!(rowname %in% c("log-posterior","mean_PPD","(Intercept)","sigma"))) %>%
  mutate(rowname=gsub("TRUE","",rowname),
         OR=exp(`50%`),
         OR_low=exp(`2.5%`),
         OR_high=exp(`97.5%`),
         type="LASSO") %>%
  arrange(-abs(`50%`))

# horseshoe prior

globalscale = (14/320)/sqrt(334)

m1_hs = stan_glm(posterior~.,
                 family=gaussian(),
                 prior=hs(global_scale=globalscale),
                 data=medication_logit)

summary(m1_hs)
s1_hs = summary(m1_hs,probs = c(0.025,.5,0.975)) %>%
  as.data.frame() %>%
  rownames_to_column() %>% 
  as_tibble() %>%
  filter(!(rowname %in% c("log-posterior","mean_PPD","(Intercept)","sigma"))) %>%
  mutate(rowname=gsub("TRUE","",rowname),
         OR=exp(`50%`),
         OR_low=exp(`2.5%`),
         OR_high=exp(`97.5%`),
         type="Regularized horseshoe") %>%
  arrange(-abs(`50%`))

post_medication = bind_rows(s1_lasso,s1_hs,s1_lin) %>%
  mutate(rowname=factor(rowname,levels=s1_hs$rowname[order(s1_hs$`50%`)]),
         type=factor(type,levels=c("Uninformative prior","LASSO","Regularized horseshoe")))
saveRDS(post_medication,file="post_medication.rds")

post_medication = readRDS("post_medication.rds")
ranks = filter(post_medication,type=="Regularized horseshoe") %>%
  arrange(-`50%`) %>%
  mutate(rank=row_number()) %>%
  mutate(medication=rowname) %>%
  dplyr::select(medication,rank)

tibble_post_medication <- post_medication %>%
  mutate(OR=paste0(sprintf("%.2f",OR)," (",sprintf("%.2f",OR_low),"-",sprintf("%.2f",OR_high),")")) %>%
  dplyr::select(medication=rowname,OR,type) %>%
  pivot_wider(names_from=type,values_from=OR) %>%
  left_join(ranks) %>%
  arrange(rank) %>% 
  filter(rank %in% c(1:10,(max(rank)-9):max(rank))) %>%
  dplyr::select(rank,medication,`Uninformative prior`,LASSO,`Regularized horseshoe`)
write.csv(tibble_post_medication,"tibble_post_medication_ranktable.csv")

post_medication %>%
  mutate(type=factor(type,labels=c("Unregularized","LASSO","Regularized Horseshoe"))) %>%
  ggplot() +
  geom_pointrange(aes(x=rowname,y=OR,ymin=OR_low,ymax=OR_high,colour=type)) +
  geom_hline(yintercept=1,linetype=2) +
  scale_colour_discrete(guide='none') +
  scale_y_log10() +
  coord_flip() +
  facet_wrap(~type,ncol=3) +
  theme(axis.text.y = element_text(size = 1)) +
  labs(y="Odds ratio (95% credible interval)",x="Medication, sex, or age",title="Association between medication and -logit(EC50) values against nAra h 2.")

write.csv(post_medication,"post_medication_ranktable.csv")

post_medication_top10 %>%
  mutate(type) %>%
  ggplot() +
  geom_pointrange(aes(reorder(x=rowname, -rank),y=OR,ymin=OR_low,ymax=OR_high,colour=type)) +
  geom_hline(yintercept=1,linetype=2) +
  scale_colour_discrete(guide='none') +
  scale_y_log10() +
  coord_flip() +
  facet_wrap(~type,ncol=3) +
  theme(axis.text.y = element_text(size = 10)) +
  labs(y="Odds ratio (95% credible interval)",x="Medication, sex, or age",title="Association between medication and -logit(EC50) values against nAra h 2.")

################################################################################
# Figs. 2F and 2H
################################################################################

ggplot(data=df_ICD10, aes(x=reorder(Var1, -Comment), y=Freq_percent, fill=Condition)) +
  geom_bar(stat="identity", position="identity", alpha = .2) +
  labs(y="Percent of condition", x='ICD-10 codes') +
  scale_fill_manual(values=c("#9e34eb", "#34baeb")) +
  coord_flip()

ggplot(data=df_medication, aes(x=reorder(Var1, -Comment), y=Freq_percent, fill=Condition)) +
  geom_bar(stat="identity", position="identity", alpha = .2) +
  labs(y="Percent of condition", x='Medications') +
  scale_fill_manual(values=c("#9e34eb", "#34baeb")) +
  coord_flip()

################################################################################
# Figs. 2J and 2K
################################################################################

#### UMAP all ICD-10 with sex, binary (cosine) metric

rnames_hits <- data_hits[,2]
print(rnames_hits)

groups <- as.factor(data_hits$HasCF[1:102])
label <- unique(groups)
print(groups)
print(label)

custom.config = umap.defaults
custom.config$metric = 'cosine'
custom.config

set.seed(42)
umap_08 <- umap::umap(data.matrix(data_hits[c(4:7387)]), metric = 'cosine')

df_umap_08 <- data.frame(x = umap_08$layout[,1],
                         y = umap_08$layout[,2],
                         color = groups)

ggplot(df_umap_08, aes(x, y, color = groups)) +
  geom_point(size = 2, stroke = 0) +
  theme_bw() + 
  ylab("UMAP Dimension 2") + 
  xlab("UMAP Dimension 1") + 
  scale_color_manual(values=c("#6A8ED2", "red")) +
  labs(colour="Has CF") +
  ggtitle('Cosine UMAP of hits: ICD-10, medication and sex')

#### PCA all ICD-10 with sex

mat_data_hits <- data.matrix(data_hits[c(4:7387)])
print(mat_data_hits)
rownames(mat_data_hits) <- rnames_hits          
print(rownames(mat_data_hits))

groups <- as.factor(data_hits$HasCF[1:102])
print(groups)

set.seed(42)
mat_data_hits.pca <- prcomp(mat_data_hits, center = TRUE)
print(mat_data_hits.pca)

fviz_eig(mat_data_hits.pca)

summary(mat_data_hits.pca)

fviz_pca_biplot(
  mat_data_hits.pca,
  axes = c(1, 2),
  geom.var = c("point", "text"),
  col.var = "grey",
  col.ind = groups,
  palette = c("#6A8ED2", "red"),
  label = "all",
  invisible = "none",
  repel = TRUE,
  habillage = "none",
  addEllipses = FALSE,
  title = "PCA of hits: ICD-10, medication and sex - show features",
  legend.title = "Has CF"
)

```

# ANALYSES PERTAINING TO FIGURE 3

```{r ANALYSES PERTAINING TO FIGURE 3, message=FALSE}

################################################################################
# Fig. 3A
################################################################################

mat_data_CF <- data.matrix(data_CF[c(3:1534, 1536)])
print(mat_data_CF)
rownames(mat_data_CF) <- rnames_CF          
print(rownames(mat_data_CF))

res.hc_binary_CF <- mat_data_CF %>%
  dist(method = "binary") %>%
  hclust(method = "ward.D2")  

fviz_dend(res.hc_binary_CF, k = 3,
          cex = 0.5,
          k_colors = c("#2E9FDF", "#FC4E07", "#E7B800"),
          color_labels_by_k = TRUE,
          rect = TRUE,
          label_cols = groups,
          show_labels = TRUE)

################################################################################
# Figs. 3B and 3C
################################################################################

#### PCA for dataset with INCLUSION of male/female sex

groups <- as.factor(data_CF$Class[1:169])
print(groups)

set.seed(42)
mat_data_CF.pca <- prcomp(mat_data_CF, center = TRUE)
print(mat_data_CF.pca)

fviz_eig(mat_data_CF.pca)

summary(mat_data_CF.pca)

fviz_pca_ind(mat_data_CF.pca,
             col.ind = groups, # color by groups
             palette = c("#6A8ED2",  "red"), 
             addEllipses = FALSE, # Concentration ellipses,
             #ellipse.type = "concentrated"
             legend.title = "Hits", 
             repel = TRUE, 
             geom.ind = c("point"),
             #geom = "points"
             title = 'PCA of CF patients: ICD-10 and sex')

#### UMAP for dataset with INCLUSION of male/female sex

groups <- as.factor(data_CF$Class[1:169])
label <- unique(groups)
print(groups)
print(label)

custom.config = umap.defaults
custom.config$metric = 'cosine'
custom.config

umap_06 <- umap::umap(data.matrix(data_CF[c(3:1534, 1536)]), metric = 'cosine')

df_umap_06 <- data.frame(x = umap_06$layout[,1],
                         y = umap_06$layout[,2],
                         color = groups)

ggplot(df_umap_06, aes(x, y, color = groups)) +
  geom_point(size = 2, stroke = 0) +
  theme_bw() + 
  ylab("UMAP Dimension 2") + 
  xlab("UMAP Dimension 1") + 
  scale_color_manual(values=c("#6A8ED2", "red")) +
  labs(colour="Hits") +
  ggtitle('Cosine UMAP of CF patients: ICD-10 and sex')

################################################################################
# Figs. 3D and 3E
################################################################################

data_sex <- read.csv("df_analysis_sex_age_patients_for_analysis_sex.csv")
rnames <- data_sex[,1]
print(rnames)

ggplot(data_sex, aes(fill=Sex, y=Number_percent, x=Condition)) + 
  geom_bar(position="stack", stat="identity")

data_age_points <- read.csv("df_analysis_sex_age_patients_for_analysis_age_points.csv")
rnames <- data_age_points[,1]
print(rnames)

ggplot(data_age_points, aes(fill=Sex, y=Age, x=Condition)) + 
  geom_boxplot()+
  geom_dotplot(position="dodge", stat="identity", binaxis='y', stackdir = 'center', dotsize = 0.1, binwidth = 1)

################################################################################
# Fig. 3F
################################################################################

Only_CF <- Only_CF %>%
  dplyr::mutate(PatientIDList = Sample_ID) %>%
  dplyr::select(-Age,-Sex, -Sample_ID)

Lab_merge <- For_merge_clean %>%
  left_join(Lab_values, by='pat_research_id') %>%
  dplyr::select(pat_research_id, PatientIDList, corrEC50, Age, Gender, observed_value:observation_value_code) %>%
  right_join(Only_CF, by='PatientIDList') %>%
  dplyr::select(pat_research_id, PatientIDList, corrEC50, Age, Gender, observed_value:observation_value_code, Condition) %>%
  dplyr::filter(observation_value_code =='186'|
                observation_value_code =='2281' |
                observation_value_code =='2278') %>%
  mutate(Lab_date = as.Date(ymd_hms(observation_date))) %>%
  mutate(J_date = substring(PatientIDList, 2,6)) %>%
  mutate(J_date = paste0('201',J_date)) %>%
  mutate(J_date = as.Date(ymd(J_date))) %>%
  rename(Unit = observed_value_unit) %>%
  dplyr::filter(J_date==Lab_date) %>% # We end up with the lab values that correspond with the day of sample processing
  mutate(across(c(contains('observed_value')), 
                ~ifelse(str_detect(., "<"), 0, .))) %>%
  mutate_at(vars(contains('observed_value')),as.numeric)

Lab_merge %>%
  ggplot(aes(x=Condition, y=observed_value, colour=Condition)) +
  geom_boxplot(outlier.shape=NA, alpha=0.7, colour='black') +
  geom_point(position = position_jitterdodge()) +
  ggh4x::facet_grid2(cols=vars(analysis_description), scales='free', independent = "y", space='free_x') +
  xlab("") +
  ylab("Respective value in g/L or mg/l") +
  scale_colour_manual(values=c('red','blue')) +
  guides(colour=FALSE) +
  theme_bw() +
  theme(strip.text = element_text(size = 12)) +
  geom_hline(data = References, aes(yintercept = (observed_value)),linetype="dashed", color = "black")
  
compare_means(observed_value ~ Condition,  data=Lab_merge, group.by = 'analysis_description',
              method = "wilcox", symnum.args=signifiance_stars, p.adjust.method = "BH")

```

# ANALYSES PERTAINING TO FIGURE 4

```{r ANALYSES PERTAINING TO FIGURE 4, message=FALSE}

################################################################################
# Figs. 4A-D
################################################################################

#### nAra h 2-specific immunoglobulins G and A

rnames <- data[,1]
print(rnames)
rownames(data) <- rnames

x <- data.matrix(data[,2:4])
rownames(x) <- rnames

my_palette <- colorRampPalette(brewer.pal(9, "Purples"))(n = 99)

colors = c(seq(1.7,4.0,length=100)) 

heatmap.2(x,
          Rowv=FALSE,
          Colv="NA",
          margin=c(8,12),
          dendrogram=c("none"),
          col= my_palette,
          symm=F,symkey=F,symbreaks=T, scale="none",
          breaks=colors,
          sepwidth=c(0.01,0.01),
          sepcolor="white",
          colsep=1:ncol(x),
          labCol=NULL,
          cexRow=1.5, cexCol=1.5,
          cellnote=labels_ige2, notecol="black", notecex=1.0,
          key=TRUE,keysize=1.5,trace="none",density.info=c("none"))

#### peanut extract ImmunoCAP

# Values

rnames <- data[,1]
print(rnames)
rownames(data) <- rnames

x <- data.matrix(data[,2:3])
rownames(x) <- rnames

my_palette <- colorRampPalette(brewer.pal(9, "Greens"))(n = 199)

colors = c(seq(0.0,200.0,length=200)) 

heatmap.2(x,
          Rowv=FALSE,
          Colv="NA",
          margin=c(8,25),
          dendrogram=c("none"),
          col= my_palette,
          symm=F,symkey=F,symbreaks=T, scale="none",
          breaks=colors,
          sepwidth=c(0.01,0.01),
          sepcolor="white",
          colsep=1:ncol(x),
          labCol=NULL,
          cexRow=0.1, cexCol=0.5,
          cellnote=labels_ige2, notecol="black", notecex=1.0,
          key=TRUE,keysize=1.5,trace="none",density.info=c("none"))

# classes

my_palette <- colorRampPalette(brewer.pal(9, "Oranges"))(n = 199)

colors = c(seq(0.0,6.0,length=200)) 

heatmap.2(x,
          Rowv=FALSE,
          Colv="NA",
          margin=c(8,25),
          dendrogram=c("none"),
          col= my_palette,
          symm=F,symkey=F,symbreaks=T, scale="none",
          breaks=colors,
          sepwidth=c(0.01,0.01),
          sepcolor="white",
          colsep=1:ncol(x),
          labCol=NULL,
          cexRow=0.1, cexCol=0.5,
          cellnote=labels_ige2, notecol="black", notecex=1.0,
          key=TRUE,keysize=1.5,trace="none",density.info=c("none"))

#### nAra h 2-specific immunoglobulins G subclasses

rnames <- data[,1]
print(rnames)
rownames(data) <- rnames

x <- data.matrix(data[,2:5])  
rownames(x) <- rnames
my_palette <- colorRampPalette(brewer.pal(9, "Purples"))(n = 99)

colors = c(seq(1.7,4.0,length=100)) 

heatmap.2(x,
          Rowv=FALSE,
          Colv="NA",
          margin=c(8,12),
          dendrogram=c("none"),
          col= my_palette,
          symm=F,symkey=F,symbreaks=T, scale="none",
          breaks=colors,
          sepwidth=c(0.01,0.01),
          sepcolor="white",
          colsep=1:ncol(x),
          labCol=NULL,
          cexRow=1.5, cexCol=1.5,
          cellnote=labels_ige2, notecol="black", notecex=1.0,
          key=TRUE,keysize=1.5,trace="none",density.info=c("none"))

#### Total immunoglobulins

data_wide <- data %>%
  select(Pat_ID, Assay, Result_corr) %>%
  pivot_wider(names_from = Assay, values_from = Result_corr) %>%
  mutate(across(-Pat_ID, ~na_if(as.character(.x), "NULL"))) %>%
  mutate(across(-Pat_ID, as.numeric)) %>%
  as.data.frame()

x_gam  <- data_wide %>% select(IgG, IgA, IgM) %>% as.matrix()
x_ige  <- data_wide %>% select(IgE) %>% as.matrix()
x_ige2 <- cbind(IgE = x_ige, IgE_copy = x_ige, IgE_copy_copy = x_ige)   # triplicate for heatmap.2
x_iggx <- data_wide %>% select(IgG1, IgG2, IgG3, IgG4) %>% as.matrix()

rownames(x_gam)  <- data_wide$Pat_ID
rownames(x_ige2) <- data_wide$Pat_ID
rownames(x_iggx) <- data_wide$Pat_ID

make_labels <- function(mat) {
  apply(mat, c(1,2), function(v) {
    if (is.na(v)) {
      return("NA")
    } else if (v == 0) {
      return("nd")
    } else {
      return(format(round(v,1), nsmall=1))
    }
  })
}

labels_gam  <- make_labels(x_gam)
labels_ige2 <- make_labels(x_ige2)
labels_iggx <- make_labels(x_iggx)

pal_blue <- colorRampPalette(brewer.pal(9, "Blues"))(99)
pal_grey <- colorRampPalette(brewer.pal(9, "Greys"))(99)

# 1) IgG, IgA, IgM (0–20 g/L)
heatmap.2(x_gam,
          Rowv=FALSE, Colv="NA", dendrogram="none",
          margin=c(8,12),
          col=pal_blue,
          symm=FALSE, symkey=FALSE, symbreaks=TRUE, scale="none",
          breaks=seq(0, 20, length=100),
          sepwidth=c(0.01,0.01), sepcolor="white", colsep=1:ncol(x_gam),
          labCol=NULL,
          cexRow=1.2, cexCol=1.5,
          key=TRUE, keysize=1.5, trace="none", density.info="none",
          main="IgG, IgA, IgM (0–20 g/L)",
          cellnote=labels_gam, notecol="black", notecex=1.0)

# 2) IgE (0–100 IU/L)
heatmap.2(x_ige2,
          Rowv=FALSE, Colv="NA", dendrogram="none",
          margin=c(8,12),
          col=pal_grey,
          symm=FALSE, symkey=FALSE, symbreaks=TRUE, scale="none",
          breaks=seq(0, 100, length=100),
          sepwidth=c(0.01,0.01), sepcolor="white", colsep=1:ncol(x_ige2),
          labCol=NULL,
          cexRow=1.2, cexCol=1.5,
          key=TRUE, keysize=1.5, trace="none", density.info="none",
          main="IgE (0–100 IU/L)",
          cellnote=labels_ige2, notecol="black", notecex=1.0)

# 3) IgG subclasses (0–20 g/L)
heatmap.2(x_iggx,
          Rowv=FALSE, Colv="NA", dendrogram="none",
          margin=c(8,12),
          col=pal_blue,
          symm=FALSE, symkey=FALSE, symbreaks=TRUE, scale="none",
          breaks=seq(0, 20, length=100),
          sepwidth=c(0.01,0.01), sepcolor="white", colsep=1:ncol(x_iggx),
          labCol=NULL,
          cexRow=1.2, cexCol=1.5,
          key=TRUE, keysize=1.5, trace="none", density.info="none",
          main="IgG Subclasses (0–20 g/L)",
          cellnote=labels_iggx, notecol="black", notecex=1.0)

################################################################################
# Fig. 4G
################################################################################

if (!all(c("Group", "ID", "Conc", "CD63") %in% colnames(data))) {
  stop("Input data must contain: Group, ID, Conc, CD63")
}

data$Group <- as.factor(data$Group)

summary_data <- data %>%
  group_by(Group, ID, Conc) %>%
  summarise(
    mean_CD63 = mean(CD63, na.rm = TRUE),
    sd_CD63   = sd(CD63, na.rm = TRUE),
    .groups = "drop"
  )

logistic_fits_id <- summary_data %>%
  group_by(Group, ID) %>%
  group_map(~ tryCatch(
    drm(mean_CD63 ~ Conc, data = .x,
        fct = LL.4(names = c("Slope","Lower","Upper","EC50"),
                   fixed = c(NA, 0, NA, NA))),
    error = function(e) NULL
  ))

id_labels <- summary_data %>% distinct(Group, ID)

pred_id <- map2_dfr(logistic_fits_id, id_labels$ID, function(mod, id) {
  if (is.null(mod)) return(NULL)
  newdata <- data.frame(Conc = 10^seq(-4, 1, length.out = 200))
  preds <- predict(mod, newdata)
  newdata$fit <- preds
  newdata$ID <- id
  return(newdata)
})

logistic_fits_group <- summary_data %>%
  group_by(Group) %>%
  group_map(~ tryCatch(
    drm(mean_CD63 ~ Conc, data = .x,
        fct = LL.4(names = c("Slope","Lower","Upper","EC50"),
                   fixed = c(NA, 0, NA, NA))),
    error = function(e) NULL
  ))

group_labels <- summary_data %>% distinct(Group)

pred_group <- map2_dfr(logistic_fits_group, group_labels$Group, function(mod, grp) {
  if (is.null(mod)) return(NULL)
  newdata <- data.frame(Conc = 10^seq(-4, 1, length.out = 200))
  preds <- predict(mod, newdata)
  newdata$fit <- preds
  newdata$Group <- grp
  return(newdata)
})

combined_model <- drm(CD63 ~ Conc,
                      curveid = Group,
                      data = data,
                      fct = LL.4(
                        names = c("Slope","Lower","Upper","EC50"),
                        fixed = c(NA, 0, NA, NA)
                      ))

ec50_mat <- as.matrix(ED(combined_model, 50, interval = "delta"))

ec50_df <- data.frame(
  Group = sub("e:(.*):50", "\\1", rownames(ec50_mat)),
  EC50 = ec50_mat[, "Estimate"],
  SE = ec50_mat[, "Std. Error"],
  Lower = ec50_mat[, "Lower"],
  Upper = ec50_mat[, "Upper"]
)

ec50_df$Lower[ec50_df$Lower < 0] <- NA

ec50_df$logEC50 <- log(ec50_df$EC50)

wald_test <- function(g1, g2, df) {
  
  row1 <- df[df$Group == g1, , drop = FALSE]
  row2 <- df[df$Group == g2, , drop = FALSE]
  
  if (nrow(row1) != 1 | nrow(row2) != 1) {
    stop(paste("Group not found:", g1, g2))
  }
  
  diff <- log(row1$EC50) - log(row2$EC50)
  
  se <- sqrt((row1$SE / row1$EC50)^2 +
               (row2$SE / row2$EC50)^2)
  
  z <- as.numeric(diff / se)
  p <- 2 * (1 - pnorm(abs(z)))
  
  data.frame(
    Comparison = paste(g1, "vs", g2),
    log_diff = as.numeric(diff),
    z_value = z,
    p_value = p
  )
}

test_CF_NC <- wald_test("CF", "NC", ec50_df)
test_Control_NC <- wald_test("Control", "NC", ec50_df)

ec50_tests <- rbind(test_CF_NC, test_Control_NC)

print(ec50_tests)

groups <- unique(summary_data$Group)

group_colors <- c(
  "CF" = "#C8559F",
  "Control" = "#374EA1",
  "NC" = "grey"
)

color_map <- c()
for (g in groups) {
  ids_in_group <- unique(summary_data$ID[summary_data$Group == g])
  base_col <- group_colors[g]
  shades <- colorRampPalette(c("grey90", base_col))(length(ids_in_group)+1)[-1]
  names(shades) <- ids_in_group
  color_map <- c(color_map, shades)
}

p <- ggplot(summary_data, aes(x = Conc, y = mean_CD63, color = ID, group = ID)) +
  geom_point(alpha = 0.7) +
  geom_errorbar(aes(ymin = mean_CD63 - sd_CD63, ymax = mean_CD63 + sd_CD63),
                width = 0.1, alpha = 0.7) +
  geom_line(data = pred_id, aes(x = Conc, y = fit, color = ID),
            size = 1, linetype = "dashed", inherit.aes = FALSE) +
  geom_line(data = pred_group, aes(x = Conc, y = fit, color = Group),
            size = 1.5, inherit.aes = FALSE) +
  geom_vline(data = ec50_df,
             aes(xintercept = EC50, color = Group),
             linetype = "dotted", size = 1, inherit.aes = FALSE) +
  geom_text(data = ec50_df,
            aes(x = EC50, y = max(summary_data$mean_CD63),
                label = paste0("EC50=", formatC(EC50, format = "g", digits = 2))),
            angle = 90, vjust = -0.5, hjust = 0, size = 3,
            inherit.aes = FALSE) +
  scale_x_log10(
    breaks = c(0.0001, 0.001, 0.01, 0.1, 1, 10),
    labels = c("0.0001", "0.001", "0.01", "0.1", "1", "10")
  ) +
  scale_color_manual(values = c(color_map, group_colors)) +
  theme_bw(base_size = 14) +
  labs(
    x = "Ara-h2 concentration (µg/mL, log scale)",
    y = "% CD63+ Basophils"
  )

print(p)

```

# ANALYSES PERTAINING TO FIGURE 5 AND S2

```{r ANALYSES PERTAINING TO FIGURE 5 AND S2, message=FALSE}

################################################################################
# Fig. 5A
################################################################################

ggplot(data=df_frequencies_all_groups, aes(reorder(x=Var1, Freq_relevant), y=Freq_percent, fill=Condition)) +
  geom_bar(stat="identity", position="dodge", alpha = .8) +
  labs(y="Percent positives within condition", x='Antigen') +
  scale_fill_manual(values=c("#0D141F", "#464849", "#B1B4B8", "#D0D2D4")) +
  theme_minimal() +
  coord_flip()

################################################################################
# Fig. 5B
################################################################################

rnames <- data[,3]
print(rnames)
rownames(data) <- rnames

x <- data.matrix(data[,4:41])
rownames(x) <- rnames

my_palette <- colorRampPalette(brewer.pal(9, "Purples"))(n = 99)

colors = c(seq(0.0,3.0,length=100)) 

heatmap.2(x,
          Rowv=FALSE,
          Colv="NA",
          margin=c(8,25),
          dendrogram=c("none"),
          col= my_palette,
          symm=F,symkey=F,symbreaks=T, scale="none",
          breaks=colors,
          sepwidth=c(0.01,0.01),
          sepcolor="white",
          colsep=1:ncol(x),
          labCol=NULL,
          cexRow=0.1, cexCol=0.5,
          key=TRUE,keysize=1.5,trace="none",density.info=c("none"))

################################################################################
# Fig. S2A-D
################################################################################

#### ELISA Heatmap

rnames <- data[,1]
print(rnames)
rownames(data) <- rnames

x <- data.matrix(data[,3:16])  # transform column 2-6 into a matrix
rownames(x) <- rnames

my_palette <- colorRampPalette(brewer.pal(9, "Purples"))(n = 99)

colors = c(seq(2.0,3.0,length=100)) 

heatmap.2(x,
          Rowv=FALSE,
          Colv="NA",
          margin=c(8,25),
          dendrogram=c("none"),
          col= my_palette,
          symm=F,symkey=F,symbreaks=T, scale="none",
          breaks=colors,
          sepwidth=c(0.01,0.01),
          sepcolor="white",
          colsep=1:ncol(x),
          cexRow=0.1, cexCol=0.5,
          key=TRUE,keysize=1.5,trace="none",density.info=c("none"))

#### ELISA PCA

rnames <- data_select[,1]
print(rnames)
rownames(data_select) <- rnames

groups <- as.factor(data_select$Cohort_type[1:933])
print(groups)

mat_data_select <- data.matrix(data_select[,3:16])
print(mat_data_select)
rownames(mat_data_select) <- rnames                  
print(rownames(mat_data_select))

set.seed(42)
mat_data_select.pca <- prcomp(mat_data_select)
print(mat_data_select.pca)

print(mat_data_select.pca$x)
write.csv(mat_data_select.pca$x, 'PCA_prcomp_output.csv')

data_violin <- read.csv("Analyses_output/PCA_prcomp_output_for_density.csv")
p <- ggplot(data_violin, aes(x=Group, y=PC2, color=Group)) + 
  geom_violin(trim=FALSE) + 
  scale_color_manual(values=c("brown","purple","orchid","blue")) +
  geom_boxplot(width=0.1) +
  geom_dotplot(binaxis='y', stackdir='center', dotsize=0.1, binwidth = 0.01) + 
  theme(legend.position="none") + 
  scale_x_discrete(limits=c("CF", "CF_paediatric", "Random_patients", "Blood_donors")) +
  ylim(-2, 3)

p

fviz_eig(mat_data_select.pca)

summary(mat_data_select.pca)

fviz_pca_biplot(
  mat_data_select.pca,
  axes = c(1, 2),
  geom.var = c("point", "text"),
  col.ind = groups,
  palette = c("red","purple","orchid","blue"), 
  col.var = "black",
  label = "var",
  invisible = "none",
  repel = TRUE,
  habillage = "none",
  addEllipses = TRUE,
  ellipse.level=0.95,
  title = "PCA - Biplot",
  labelsize = 4
)

#### ELISA HC

dend <- mat_data_select %>% dist %>% hclust %>% as.dendrogram
dend_info <- mat_data_select %>% dist %>% hclust
dend_info_outpout <- data.frame(dendextend::cutree(dend_info,2))
write.csv(dend_info_outpout, 'dend_info_outpout.csv')

dend <- dend %>% 
  set("labels_colors", as.numeric(groups), order_value = TRUE) %>%
  set("labels_cex", .25) %>%
  set("branches_k_color", k=2)
par(mar = c(4,1,0,8))
circlize_dendrogram(dend)

#### ISAC Heatmaps

# IgG4 data without cutoff

rnames <- data_IgG4[,3]
print(rnames)
rownames(data_IgG4) <- rnames

x <- data.matrix(data_IgG4[,4:22])
rownames(x) <- rnames

my_palette <- colorRampPalette(brewer.pal(9, "Purples"))(n = 99)

colors = c(seq(0.0,2.0,length=100)) 

heatmap.2(x,
          Rowv=FALSE,
          Colv="NA",
          margin=c(8,25),
          dendrogram=c("none"),
          col= my_palette,
          symm=F,symkey=F,symbreaks=T, scale="none",
          breaks=colors,
          sepwidth=c(0.01,0.01),
          sepcolor="white",
          colsep=1:ncol(x),
          labCol=NULL,
          cexRow=0.1, cexCol=0.5,
          key=TRUE,keysize=1.5,trace="none",density.info=c("none"))

# IgE data without cutoff

rnames <- data_IgE[,3]
print(rnames)
rownames(data_IgG4) <- rnames

x <- data.matrix(data_IgE[,4:22])
rownames(x) <- rnames

my_palette <- colorRampPalette(brewer.pal(9, "Purples"))(n = 99)

colors = c(seq(0.0,85,length=100)) 

heatmap.2(x,
          Rowv=FALSE,
          Colv="NA",
          margin=c(8,25),
          dendrogram=c("none"),
          col= my_palette,
          symm=F,symkey=F,symbreaks=T, scale="none",
          breaks=colors,
          sepwidth=c(0.01,0.01),
          sepcolor="white",
          colsep=1:ncol(x),
          labCol=NULL,
          cexRow=0.1, cexCol=0.5,
          key=TRUE,keysize=1.5,trace="none",density.info=c("none"))

```

# ANALYSES PERTAINING TO FIGURE 6

```{r ANALYSES PERTAINING TO FIGURE 6, message=FALSE}

################################################################################
# Fig. 6A
################################################################################

rnames <- data[,1]
print(rnames)
rownames(data) <- rnames

x <- data.matrix(data[,2:56])
rownames(x) <- rnames

my_palette <- colorRampPalette(brewer.pal(9, "Purples"))(n = 99)

colors = c(seq(0.0,10.0,length=100)) 

heatmap.2(x,
          Rowv=FALSE,
          Colv="NA",
          margin=c(8,25),dendrogram=c("none"),
          col= my_palette,
          symm=F,symkey=F,symbreaks=T, scale="none",
          breaks=colors,
          sepwidth=c(0.01,0.01),
          sepcolor="white",
          colsep=1:ncol(x),
          labCol=NULL,
          cexRow=0.005, cexCol=0.75,
          key=TRUE,keysize=1.5,trace="none",density.info=c("none"))

################################################################################
# Fig. 6B
################################################################################

rnames <- data_select[,1]
print(rnames)
rownames(data_select) <- rnames

mat_data_select <- data.matrix(data_select[,2:56])
print(mat_data_select)
rownames(mat_data_select) <- rnames                  
print(rownames(mat_data_select))

set.seed(42)
mat_data_select.pca <- prcomp(mat_data_select, center = TRUE)
print(mat_data_select.pca)

fviz_eig(mat_data_select.pca)

summary(mat_data_select.pca)

fviz_pca_biplot(
  mat_data_select.pca,
  axes = c(1, 2),
  geom.var = c("point", "text"),
  col.ind = "cos2",
  fill.ind = "white",
  col.var = "black",
  fill.var = "white",
  gradient.cols = "RdBu",
  label = "var",
  invisible = "none",
  repel = TRUE,
  habillage = "none",
  palette = "RdBu",
  addEllipses = FALSE,
  title = "PCA - Biplot",
  labelsize = 1
)

################################################################################
# Fig. 6C
################################################################################

rnames <- data[,1]
print(rnames)

volcano <- data.frame(data[c(2:3)])

options(ggrepel.max.overlaps = Inf)

ggplot(volcano, aes(x=CF.CTRL, y=(Average_all), label=data$Protein)) +
  geom_point(size = 2, stroke = 0) +
  geom_label_repel(aes(label=ifelse(CF.CTRL>1 | CF.CTRL<(-1) & Average_all>1, as.character(data$Protein),''))) +
  geom_hline(yintercept = 1, 
             colour="#990000", 
             linetype="dashed") + 
  geom_vline(xintercept = 1, 
             colour="#990000", 
             linetype="dashed") + 
  geom_vline(xintercept = -1, 
             colour="#990000", 
             linetype="dashed") +
  theme_bw(base_size = 14) +
  xlim(-2.8, 2.8) +
  ylab("Average log2 fold change") + 
  xlab("Log2 fold change (CF-CTRL)") 

```

# ANALYSES PERTAINING TO FIGURE 7

```{r ANALYSES PERTAINING TO FIGURE 7, message=FALSE}

################################################################################
# Fig. 7D
################################################################################

data_HLA_long %>%
  dplyr::filter(Patient_ID=='USZ-HTA-020' | Patient_ID=='USZ-HTA-014') %>%
  dplyr::filter(Cytokine =='IL-4' | Cytokine=='TNF-alpha')  %>%
  dplyr::filter(!(Protein =='Whole peanut extract'| Protein=='Ara H2  whole protein'))  %>%
  ggplot(aes(x = reorder(Protein,Order), y = Signal, color = Patient_ID, fill=Patient_ID, group=Patient_ID)) +
  geom_point(position='dodge', size=5) +
  geom_line() +
  xlab("") +
  ylab("Signal") +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  ggh4x::facet_grid2(rows=vars(Cytokine), cols= vars(Sample_type), scales='free', independent = "y") +
  force_panelsizes(cols = c(0.1, 0.2, 1.1)) +
  scale_colour_viridis_d() +
  scale_fill_viridis_d() +
  theme_bw()

```

# ANALYSES PERTAINING TO FIGURE 8

```{r ANALYSES PERTAINING TO FIGURE 8, message=FALSE}

################################################################################
# Fig. 8E and 8F
################################################################################

col <- brewer.pal(n=11, name="RdBu")
col2 <- c('#C7549F', '#374DA1')
render.cont <- function(x) { with(stats.default(x), c("", "Mean (SD)" = sprintf("%0.1f (%0.1f)", MEAN, SD), "Median (IQR)" = sprintf("%0.1f (%0.1f to %0.1f)", MEDIAN, Q1, Q3), "Range" = sprintf("%0.0f to %0.0f", MIN, MAX))) }
render.cont.eq5d <- function(x) { with(stats.default(x), c("", "Mean (SD)" = sprintf("%0.2f (%0.2f)", MEAN, SD), "Median (IQR)" = sprintf("%0.2f (%0.2f to %0.2f)", MEDIAN, Q1, Q3), "Range" = sprintf("%0.2f to %0.2f", MIN, MAX))) }
render.cat <- function(x) { c("", sapply(stats.default(x), function(y) with(y, sprintf("%d (%0.1f%%)", FREQ, PCTnoNA)))) }

data$Year <- as.numeric(data$Year)
data$DOB <- as.numeric(data$DOB)

data$Sex <- recode_factor(data$Sex, "1" = "female", "2" = "male", "3" = "neither female nor male")
data$Sex <- as.factor(data$Sex)

data = data %>% 
  mutate(bmi = Weight/(Height/100)^2)

data$Age <- as.numeric(data$Year-data$DOB)
data$Smoking <- recode_factor(data$Smoking, "1" = "Yes, daily", "2" = "Yes, sometimes", "3" = "Ex-smoker", "4" = "Non-smoker")
data$Alcohol <- recode_factor(data$Alcohol, "1" = "Never", "2" = "Once per month or less", "3" = "2-3 times per month", "4" = "2-3 times per week", "5" = "4 times or more per week")
data$CFTR_genotype <- recode_factor(data$CFTR_genotype, "1" = "Wildtyp", "2" = "Mutationstraeger", "3" = "Mukoviszidose")
data$Which_mut_if_carrier <- recode_factor(data$Which_mut_if_carrier, 
                                           "0" = "Ich weiss nicht genau, aber ich bin Mutationstraeger",
                                           "1" = "Ich weiss nicht genau, aber mein Kind hat Mukoviszidose", 
                                           "2" = "F508del", 
                                           "3" = "G542X",
                                           "4" = "G551D", 
                                           "5" = "N1303K", 
                                           "6" = "R117H",
                                           "7" = "W1282X", 
                                           "8" = "R553X", 
                                           "9" = "621+1G->T",
                                           "10" = "1717-1G->A", 
                                           "11" = "3849+10kbC->T", 
                                           "12" = "2789+5G->A",
                                           "13" = "I507del", 
                                           "14" = "R1162X", 
                                           "15" = "G85E",
                                           "16" = "D1152H", 
                                           "17" = "2183AA->G", 
                                           "18" = "3659delC",
                                           "19" = "R347P", 
                                           "20" = "5T", 
                                           "21" = "3120+1G->A",
                                           "22" = "A455E", 
                                           "23" = "3272-26A->G",
                                           "24" = "R334W", 
                                           "25" = "1898+1G->A",
                                           "26" = "CFTRdele2, 3", 
                                           "27" = "R560T", 
                                           "28" = "L206W",
                                           "29" = "2184insA", 
                                           "30" = "394delTT", 
                                           "31" = "E60X",
                                           "32" = "Q493X", 
                                           "33" = "Eine andere, hier nicht genannte")

data$Which_mut_if_CF <- recode_factor(data$Which_mut_if_CF, 
                                           "0" = "Ich weiss nicht genau. Aber ich habe Mukoviszidose",
                                           "1" = "Ich habe zwei verschiedene Krankheit verursachende Mutationen", 
                                           "2" = "F508del", 
                                           "3" = "G542X",
                                           "4" = "G551D", 
                                           "5" = "N1303K", 
                                           "6" = "R117H",
                                           "7" = "W1282X", 
                                           "8" = "R553X", 
                                           "9" = "621+1G->T",
                                           "10" = "1717-1G->A", 
                                           "11" = "3849+10kbC->T", 
                                           "12" = "2789+5G->A",
                                           "13" = "I507del", 
                                           "14" = "R1162X", 
                                           "15" = "G85E",
                                           "16" = "D1152H", 
                                           "17" = "2183AA->G", 
                                           "18" = "3659delC",
                                           "19" = "R347P", 
                                           "20" = "5T", 
                                           "21" = "3120+1G->A",
                                           "22" = "A455E", 
                                           "23" = "3272-26A->G",
                                           "24" = "R334W", 
                                           "25" = "1898+1G->A",
                                           "26" = "CFTRdele2, 3", 
                                           "27" = "R560T", 
                                           "28" = "L206W",
                                           "29" = "2184insA", 
                                           "30" = "394delTT", 
                                           "31" = "E60X",
                                           "32" = "Q493X", 
                                           "33" = "Eine andere, hier nicht genannte")

data$Allergy <- recode_factor(data$Allergy, "1" = "Yes", "0" = "No")
data$Cardiac <- recode_factor(data$Cardiac, "1" = "Yes", "0" = "No")
data$Stroke <- recode_factor(data$Stroke, "1" = "Yes", "0" = "No")
data$Cancer <- recode_factor(data$Cancer, "1" = "Yes", "0" = "No")
data$Depression <- recode_factor(data$Depression, "1" = "Yes", "0" = "No")
data$Diabetes_mel <- recode_factor(data$Diabetes_mel, "1" = "Yes", "0" = "No")
data$Hypertension <- recode_factor(data$Hypertension, "1" = "Yes", "0" = "No")
data$Arthritis <- recode_factor(data$Arthritis, "1" = "Yes", "0" = "No")
data$Osteoporosis <- recode_factor(data$Osteoporosis, "1" = "Yes", "0" = "No")
data$Asthma_bronchiale <- recode_factor(data$Asthma_bronchiale, "1" = "Yes", "0" = "No")
data$COPD <- recode_factor(data$COPD, "1" = "Yes", "0" = "No")
data$Hepatitis <- recode_factor(data$Hepatitis, "1" = "Yes", "0" = "No")
data$Infl_bowel <- recode_factor(data$Infl_bowel, "1" = "Yes", "0" = "No")
data$Chronic_kidney <- recode_factor(data$Chronic_kidney, "1" = "Yes", "0" = "No")
data$Autoimmune <- recode_factor(data$Autoimmune, "1" = "Yes", "0" = "No")
data$Atopic_dermatitis <- recode_factor(data$Atopic_dermatitis, "1" = "Yes", "0" = "No")
data$Migraine <- recode_factor(data$Migraine, "1" = "Yes", "0" = "No")
data$CF <- recode_factor(data$CF, "1" = "Yes", "0" = "No")
data$HIV <- recode_factor(data$HIV, "1" = "Yes", "0" = "No")
data$Other_immune <- recode_factor(data$Other_immune, "1" = "Yes", "0" = "No")

data$Pollen <- recode_factor(ifelse(data$Allergy == 'No', 'No', data$Pollen), "1" = "Yes", "0" = "No")
data$Contact <- recode_factor(ifelse(data$Allergy == 'No', 'No', data$Contact), "1" = "Yes", "0" = "No")
data$Food <- recode_factor(ifelse(data$Allergy == 'No', 'No', data$Food), "1" = "Yes", "0" = "No")
data$Other_allergy <- recode_factor(ifelse(data$Allergy == 'No', 'No', data$Other_allergy), "1" = "Yes", "0" = "No")

data$Hasel <- recode_factor(ifelse(data$Pollen == 'No', 'No', data$Hasel), "1" = "Yes", "0" = "No")
data$Erle <- recode_factor(ifelse(data$Pollen == 'No', 'No', data$Erle), "1" = "Yes", "0" = "No")
data$Esche <- recode_factor(ifelse(data$Pollen == 'No', 'No', data$Esche), "1" = "Yes", "0" = "No")
data$Birke <- recode_factor(ifelse(data$Pollen == 'No', 'No', data$Birke), "1" = "Yes", "0" = "No")
data$Hagebuche <- recode_factor(ifelse(data$Pollen == 'No', 'No', data$Hagebuche), "1" = "Yes", "0" = "No")
data$Eiche <- recode_factor(ifelse(data$Pollen == 'No', 'No', data$Eiche), "1" = "Yes", "0" = "No")
data$Graeser <- recode_factor(ifelse(data$Pollen == 'No', 'No', data$Graeser), "1" = "Yes", "0" = "No")
data$Kraeuter <- recode_factor(ifelse(data$Pollen == 'No', 'No', data$Kraeuter), "1" = "Yes", "0" = "No")
data$Other_allerg <- recode_factor(ifelse(data$Pollen == 'No', 'No', data$Other_allerg), "1" = "Yes", "0" = "No")

data$Gluten <- recode_factor(ifelse(data$Food == 'No', 'No', data$Gluten), "1" = "Yes", "0" = "No")
data$Milch <- recode_factor(ifelse(data$Food == 'No', 'No', data$Milch), "1" = "Yes", "0" = "No")
data$Eier <- recode_factor(ifelse(data$Food == 'No', 'No', data$Eier), "1" = "Yes", "0" = "No")
data$Fische <- recode_factor(ifelse(data$Food == 'No', 'No', data$Fische), "1" = "Yes", "0" = "No")
data$Krebstiere <- recode_factor(ifelse(data$Food == 'No', 'No', data$Krebstiere), "1" = "Yes", "0" = "No")
data$Soja <- recode_factor(ifelse(data$Food == 'No', 'No', data$Soja), "1" = "Yes", "0" = "No")
data$Nuesse <- recode_factor(ifelse(data$Food == 'No', 'No', data$Nuesse), "1" = "Yes", "0" = "No")
data$Erdnuesse <- recode_factor(ifelse(data$Food == 'No', 'No', data$Erdnuesse), "1" = "Yes", "0" = "No")
data$Sesam <- recode_factor(ifelse(data$Food == 'No', 'No', data$Sesam), "1" = "Yes", "0" = "No")
data$Sellerie <- recode_factor(ifelse(data$Food == 'No', 'No', data$Sellerie), "1" = "Yes", "0" = "No")
data$Senf <- recode_factor(ifelse(data$Food == 'No', 'No', data$Senf), "1" = "Yes", "0" = "No")
data$Lupinen <- recode_factor(ifelse(data$Food == 'No', 'No', data$Lupinen), "1" = "Yes", "0" = "No")
data$Weichtiere <- recode_factor(ifelse(data$Food == 'No', 'No', data$Weichtiere), "1" = "Yes", "0" = "No")
data$Other <- recode_factor(ifelse(data$Food == 'No', 'No', data$Other), "1" = "Yes", "0" = "No")

table1(~ Age + Sex | CFTR_genotype,
       render.continuous = render.cont, render.categorical = render.cat, 
       data = data)

data %>%
  filter(!CFTR_genotype=='Wildtyp') %>%
  pivot_longer(cols = Allergy:Other, names_to = "symptom", values_to = "value") %>% 
  group_by(CFTR_genotype, symptom, value) %>% 
  summarise(n = n()) %>% 
  filter(!is.na(value)) %>% 
  mutate(ntot = sum(n), freq = n / ntot) %>% 
  ungroup() %>% 
  filter(value == "Yes") %>% 
  mutate(symptom = fct_reorder(symptom, freq)) %>%
  ggplot(aes(y = symptom, x = freq, fill = fct_rev(CFTR_genotype))) + 
  geom_bar(stat = "identity", position = position_dodge(preserve = "single")) + 
  scale_fill_manual(values = col2) + 
  guides(fill = guide_legend(reverse = TRUE)) + 
  theme_bw() + 
  labs(fill = "CFTR_genotype")

data %>%
  filter(!CFTR_genotype=='Wildtyp') %>%
  pivot_longer(cols = Which_mut_if_carrier:Which_mut_if_CF, names_to = "symptom", values_to = "value") %>% 
  group_by(CFTR_genotype, symptom, value) %>% 
  summarise(n = n()) %>% 
  filter(!is.na(value)) %>% 
  mutate(ntot = sum(n), freq = n / ntot) %>% 
  ungroup() %>% 
  mutate(value = fct_reorder(value, freq)) %>%
  ggplot(aes(y = value, x = freq, fill = fct_rev(CFTR_genotype))) + 
  geom_bar(stat = "identity", position = position_dodge(preserve = "single")) + 
  scale_fill_manual(values = col2) + 
  guides(fill = guide_legend(reverse = TRUE)) + 
  theme_bw() + 
  labs(fill = "CFTR_genotype")

```

# ANALYSES PERTAINING TO FIGURE 9 AND S7

```{r ANALYSES PERTAINING TO FIGURE 9 AND S7, message=FALSE}

################################################################################
# Fig. 9B-D AND S7A-D
################################################################################

#### Body temperature and anaphylaxis

# -------------------------- AUC --------------------------

calc_auc_safe <- function(time, signal) {
  df <- data.frame(time = time, signal = signal)
  df <- df[complete.cases(df), ]
  df <- df %>%
    group_by(time) %>%
    summarise(signal = mean(signal), .groups = "drop") %>%
    arrange(time)
  if(nrow(df) < 2) return(NA_real_)
  sum(diff(df$time) * (head(df$signal, -1) + tail(df$signal, -1)) / 2)
}

auc_ana <- data_ana_long %>%
  group_by(Mouse_number, Group, Adjuvans, Type) %>%
  summarise(AUC = calc_auc_safe(Time, Signal), .groups = "drop") %>%
  filter(!is.na(AUC))

auc_temp <- data_temp_long %>%
  group_by(Mouse_number, Group, Adjuvans, Type) %>%
  summarise(AUC = calc_auc_safe(Time, Signal), .groups = "drop") %>%
  filter(!is.na(AUC))

# Bootstrapped 95% CI function
boot_ci <- function(x, R = 1000, conf = 0.95) {
  if(length(x) < 2) return(c(lower = NA, upper = NA))
  boot_samples <- replicate(R, mean(sample(x, replace = TRUE)))
  quantile(boot_samples, probs = c((1 - conf)/2, 1 - (1 - conf)/2))
}

# Prepare AUC summary with CI for plotting
auc_ana_summary <- auc_ana %>%
  group_by(Group, Adjuvans, Type) %>%
  summarise(
    mean_AUC = mean(AUC),
    lower = boot_ci(AUC)[1],
    upper = boot_ci(AUC)[2],
    .groups = "drop"
  )

auc_temp_summary <- auc_temp %>%
  group_by(Group, Adjuvans, Type) %>%
  summarise(
    mean_AUC = mean(AUC),
    lower = boot_ci(AUC)[1],
    upper = boot_ci(AUC)[2],
    .groups = "drop"
  )

# -------------------------- Custom CI function for ribbons --------------------------
ci_mean <- function(x) {
  m <- mean(x)
  se <- sd(x)/sqrt(length(x))
  data.frame(y = m, ymin = m - 1.96*se, ymax = m + 1.96*se)
}

# -------------------------- Plots --------------------------

# AUC for anaphylactic score
ggplot(auc_ana, aes(x = interaction(Group, Adjuvans, Type), y = AUC,
                    color = interaction(Group, Adjuvans, Type),
                    fill  = interaction(Group, Adjuvans, Type))) +
  geom_point(size = 3, position = position_jitter(width = 0.2)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.2, color = 'black') +
  scale_fill_manual(values = c('#0d0887', '#6a00a8', '#b12a90', '#e16462', '#f0f921', 'lightgrey')) +
  scale_colour_manual(values = c('#0d0887', '#6a00a8', '#b12a90', '#e16462', '#f0f921', 'lightgrey')) +
  theme_bw() +
  labs(y = "AUC (Anaphylactic score)", x = '')

# Individual curves + mean + CI ribbons for anaphylaxis
ggplot(data_ana_long, aes(Time, Signal, group = Mouse_number, color = Group)) +
  geom_line(alpha = 0.3) +
  stat_summary(aes(group = Group), fun = mean, geom = "line", size = 1.5) +
  stat_summary(aes(group = Group, fill = Group), fun.data = ci_mean, geom = "ribbon", alpha = 0.2, color = NA) +
  scale_fill_manual(values = c('#0d0887', '#6a00a8', '#b12a90', '#e16462', '#f0f921', 'lightgrey')) +
  scale_colour_manual(values = c('#0d0887', '#6a00a8', '#b12a90', '#e16462', '#f0f921', 'lightgrey')) +
  theme_bw() +
  labs(y = "Anaphylactic score", x = "Time (minutes)")

# AUC for temperature
ggplot(auc_temp, aes(x = interaction(Group, Adjuvans, Type), y = AUC,
                     color = interaction(Group, Adjuvans, Type),
                     fill  = interaction(Group, Adjuvans, Type))) +
  geom_point(size = 3, position = position_jitter(width = 0.2)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.2, color = 'black') +
  scale_fill_manual(values = c('#0d0887', '#6a00a8', '#b12a90', '#e16462', '#f0f921', 'lightgrey')) +
  scale_colour_manual(values = c('#0d0887', '#6a00a8', '#b12a90', '#e16462', '#f0f921', 'lightgrey')) +
  theme_bw() +
  labs(y = "AUC (Temperature)", x = '')

# Individual curves + mean + CI ribbons for temperature
ggplot(data_temp_long, aes(Time, Signal, group = Mouse_number, color = Group)) +
  geom_line(alpha = 0.3) +
  stat_summary(aes(group = Group), fun = mean, geom = "line", size = 1.5) +
  stat_summary(aes(group = Group, fill = Group), fun.data = ci_mean, geom = "ribbon", alpha = 0.2, color = NA) +
  scale_fill_manual(values = c('#0d0887', '#6a00a8', '#b12a90', '#e16462', '#f0f921', 'lightgrey')) +
  scale_colour_manual(values = c('#0d0887', '#6a00a8', '#b12a90', '#e16462', '#f0f921', 'lightgrey')) +
  theme_bw() +
  labs(y = "Body temperature", x = "Time (minutes)")

# -------------------------- Significance testing --------------------------

comparisons <- list(
  c("CF", "WT"),
  c("CF gut cr", "WT")
)

run_wilcox_auc <- function(df, group1, group2) {
  df %>%
    summarise(
      p_value = wilcox.test(
        AUC[Group == group1],
        AUC[Group == group2],
        exact = FALSE
      )$p.value
    ) %>%
    mutate(comparison = paste(group1, "vs", group2))
}

results_ana <- bind_rows(
  lapply(comparisons, function(comp)
    run_wilcox_auc(
      subset(auc_ana,
             Type == "Challenge" &
               Adjuvans == "CT" &
               Group %in% c("CF", "CF gut cr", "WT")),
      comp[1], comp[2]
    )
  )
) %>%
  mutate(p_adj = p.adjust(p_value, method = "holm")) %>%
  mutate(
    significance = case_when(
      p_adj < 0.0001 ~ "****",
      p_adj < 0.001  ~ "***",
      p_adj < 0.01   ~ "**",
      p_adj < 0.05   ~ "*",
      TRUE           ~ "ns"
    )
  )

results_ana

results_temp <- bind_rows(
  lapply(comparisons, function(comp)
    run_wilcox_auc(
      subset(auc_temp,
             Type == "Challenge" &
               Adjuvans == "CT" &
               Group %in% c("CF", "CF gut cr", "WT")),
      comp[1], comp[2]
    )
  )
) %>%
  mutate(p_adj = p.adjust(p_value, method = "holm")) %>%
  mutate(
    significance = case_when(
      p_adj < 0.0001 ~ "****",
      p_adj < 0.001  ~ "***",
      p_adj < 0.01   ~ "**",
      p_adj < 0.05   ~ "*",
      TRUE           ~ "ns"
    )
  )

results_temp

#### Cytokines

level_order <- c('Baseline', 'Immunisation', 'Challenge')

level2 <- c('Peanut', 'none')

subset(data_cyto_long,
       Cytokine %in% c('IL-4','IFN_gamma','IL-10','IL-17A','IL-13')) %>% 
  ggplot(aes(
    y = factor(Type, level = level_order),
    x = Signal,
    fill = interaction(Group, Adjuvans, factor(Treatment, level = level2)),
    colour = interaction(Group, Adjuvans, factor(Treatment, level = level2))
  )) +
  geom_boxplot(outlier.shape = NA, alpha=0.2, color='black') +
  geom_point(position = position_jitterdodge(), alpha = 1, size=2) +
  coord_flip() +
  scale_fill_manual(values = c('#0d0887', '#6a00a8','#b12a90','#e16462','#fca636','#f0f921','lightgrey'),
                    name='Group-Adjuvans-Treatment') +
  scale_colour_manual(values = c('#0d0887', '#6a00a8','#b12a90','#e16462','#fca636','#f0f921','lightgrey'),
                      name='Group-Adjuvans-Treatment') +
  ggh4x::facet_grid2(rows = vars(Cytokine),
                     cols = vars(factor(Type, level = level_order)),
                     scales = 'free', independent = "y", space = 'free_x', drop = FALSE) +
  force_panelsizes(cols = c(0.5, 1, 1)) +
  labs(y = "Before or after challenge", x = 'Plasma cytokine concentration (pg/ml)') +
  theme_bw()

# We condcut some statistical tests

# Groups to compare
comparisons <- list(
  c("CF", "WT"),
  c("CF gut cr", "WT")
)

# Focus only on CT adjuvant and Peanut treatment
data_cyto_focus <- subset(data_cyto_long,
                          Adjuvans == "CT" & Treatment == "Peanut")

# Function to run Wilcoxon per cytokine and Type
run_wilcox <- function(df, group1, group2) {
  df %>%
    group_by(Type, Cytokine) %>%
    summarise(
      p_value = wilcox.test(Signal[Group == group1],
                            Signal[Group == group2],
                            exact = FALSE)$p.value,
      .groups = "drop"
    ) %>%
    mutate(
      comparison = paste(group1, "vs", group2)
    )
}

# Run all comparisons
results <- bind_rows(
  lapply(comparisons, function(comp) run_wilcox(data_cyto_focus, comp[1], comp[2]))
)

# Adjust for multiple comparisons (FDR per cytokine)
results <- results %>%
  group_by(Cytokine) %>%
  mutate(p_adj = p.adjust(p_value, method = "fdr")) %>%
  ungroup() %>%
  mutate(
    significance = case_when(
      p_adj < 0.0001 ~ "****",
      p_adj < 0.001  ~ "***",
      p_adj < 0.01   ~ "**",
      p_adj < 0.05   ~ "*",
      TRUE           ~ "ns"
    )
  )

print(results)

#### ELISA data

# Focused ELISA data, remove 'untreated' week
data_ELISA_focus <- data_ELISA_long %>%
  filter(Week != "untreated") %>%
  mutate(
    Signal = as.numeric(Signal),
    Week = factor(Week, levels = c(2, 5, "Challenge")),
    Immunoglobulin = factor(Immunoglobulin, levels = rev(c("IgG", "IgE"))),  # IgE top, IgG bottom
    Group = factor(Group, levels = c("WT", "CF", "CF_gut"))
  )

# Define colors
colors_vec <- c('#0d0887', '#6a00a8','#b12a90','#e16462','#fca636','#f0f921','lightgrey')

# Function to plot ELISA

plot_ELISA <- function(df, log_transform = FALSE) {
  
  # Reorder Group factor
  df$Group <- factor(df$Group, levels = c("CF", "CF_gut", "WT"))
  
  # Optionally log-transform the y-axis
  y_aes <- if(log_transform) {
    log10(df$Signal)
  } else {
    df$Signal
  }
  
  ggplot(df, aes(
    y = y_aes,
    x = Group,
    fill = Group,
    colour = Group
  )) +
    geom_boxplot(outlier.shape = NA, alpha = 0.2, color='black') +
    geom_point(position = position_jitterdodge(jitter.width = 0.15), alpha = 0.8, size=2) +
    scale_fill_manual(values = c('CF'='#0d0887','CF_gut'='#6a00a8','WT'='#b12a90')) +
    scale_colour_manual(values = c('CF'='#0d0887','CF_gut'='#6a00a8','WT'='#b12a90')) +
    ggh4x::facet_grid2(
      rows = vars(Immunoglobulin),
      cols = vars(Week),
      space = "free_x",
      drop = FALSE
    ) +
    labs(
      x = "",
      y = ifelse(log_transform, "log10(AU/ml)", "Antibody AU/ml")
    ) +
    theme_bw()
}

# Plot
plot_ELISA(data_ELISA_focus, log_transform = FALSE)   # raw
plot_ELISA(data_ELISA_focus, log_transform = TRUE)    # log10

################################
# Step 1: Compute IgE/IgG ratio
################################

# Pivot wider so IgG and IgE are columns
data_ELISA_ratio <- data_ELISA_focus %>%
  pivot_wider(
    id_cols = c(Adjuvans, Treatment, Week, Group, Mouse_number, Antigen, Experiment_year), 
    names_from = Immunoglobulin, 
    values_from = Signal
  ) %>%
  mutate(
    IgG_IgE_ratio = IgG / IgE,
    log_IgG_IgE_ratio = log10(IgG / IgE)
  )

head(data_ELISA_ratio)

# Reorder Group factor
data_ELISA_ratio$Group <- factor(data_ELISA_ratio$Group, levels = c("CF", "CF_gut", "WT"))

################################
# Step 2: Plot log-ratio
################################

plot_ELISA_ratio <- function(df, use_log = TRUE) {
  y_col <- if(use_log) "log_IgG_IgE_ratio" else "IgG_IgE_ratio"
  
  ggplot(df, aes(
    y = .data[[y_col]],
    x = Group,
    fill = Group,
    colour = Group
  )) +
    geom_boxplot(outlier.shape = NA, alpha = 0.2, color='black') +
    geom_point(position = position_jitterdodge(jitter.width = 0.15), alpha = 0.8, size=2) +
    scale_fill_manual(values = c('CF'='#0d0887','CF_gut'='#6a00a8','WT'='#b12a90')) +
    scale_colour_manual(values = c('CF'='#0d0887','CF_gut'='#6a00a8','WT'='#b12a90')) +
    ggh4x::facet_grid2(
      cols = vars(Week),
      drop = FALSE
    ) +
    labs(
      x = "",
      y = ifelse(use_log, "log10(IgG / IgE ratio)", "IgG / IgE ratio")
    ) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 0, hjust = 0.5)
    )
}

# Plot
plot_ELISA_ratio(data_ELISA_ratio, use_log = TRUE)   # log-ratio
plot_ELISA_ratio(data_ELISA_ratio, use_log = FALSE)  # raw ratio

# Groups to compare
comparisons <- list(
  c("CF", "WT"),
  c("CF_gut", "WT")
)

# Run Wilcoxon for log IgG/IgE ratio per Week
run_wilcox_ratio <- function(df, group1, group2) {
  df %>%
    group_by(Week) %>%
    summarise(
      p_value = wilcox.test(log_IgG_IgE_ratio[Group == group1],
                            log_IgG_IgE_ratio[Group == group2],
                            exact = FALSE)$p.value,
      .groups = "drop"
    ) %>%
    mutate(
      comparison = paste(group1, "vs", group2)
    )
}

# Run all comparisons
results_ratio <- bind_rows(
  lapply(comparisons, function(comp) run_wilcox_ratio(data_ELISA_ratio, comp[1], comp[2]))
)

# FDR correction per week
results_ratio <- results_ratio %>%
  group_by(Week) %>%
  mutate(p_adj = p.adjust(p_value, method = "fdr")) %>%
  ungroup() %>%
  mutate(
    significance = case_when(
      p_adj < 0.0001 ~ "****",
      p_adj < 0.001  ~ "***",
      p_adj < 0.01   ~ "**",
      p_adj < 0.05   ~ "*",
      TRUE           ~ "ns"
    )
  )

results_ratio

```
